---
title: "RepData_PeerAssessment2"
author: "Gary"
date: "August 17, 2015"
output: html_document
---

##U.S. Storms: The Physical and Financial Harm (1950 - 2011)

##Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.  In this study, we strive to answer the following questions:

1. Which types of events are most harmful with respect to population health?

2. Which types of events have the greatest economic consequences?

##Data Processing


```{r}
## Load the appropriate libraries
library("dplyr")
library("ggplot2")

## Read in NoAA data - note that read.csv can read directly from compressed files
noaaDB <- read.csv("repdata-data-StormData.csv.bz2")

## For dplyr, tbl_df is easier to use
noaaDB <- tbl_df(noaaDB)

glimpse(noaaDB)

## Rank Event Types by cost to human health
dfDamage <- noaaDB %>%
    group_by(EVTYPE) %>%
    summarise(p_Fatal = sum(FATALITIES, na.rm = TRUE), p_Injured = sum(INJURIES, na.rm = TRUE)) %>%
    mutate(rank = rank(p_Fatal)) %>%
    arrange(desc(rank), desc(p_Injured))


## Rank Event Types by cost to economic health
dfDamageEcon <- noaaDB %>%
    group_by(EVTYPE) %>%
    summarise(p_PropDmg = sum(PROPDMG, na.rm = TRUE), p_CropDmg = sum(CROPDMG, na.rm = TRUE)) %>%
    mutate(rank = rank(p_PropDmg)) %>%
    arrange(desc(rank), desc(p_CropDmg))


totImpace <- full_join(dfDamage, dfDamageEcon, by = "EVTYPE")

```



##Results


```{r}
head(dfDamage)
```


```{r}
head(dfDamageEcon)
```



```{r}
head(totImpace)
```










